---
title: "Discrete Partial Credit Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Discrete Partial Credit Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CPTtools)
```

# Required Software

## Mongo Database

The Proc4 system is designed to work with the Mongo database.  This
can be downloaded and installed from https://www.mongodb.org/.  The
community edition works fine.  Assuming that the R processes are
running on the same server as the database, it does not need to be
configured to allow access from hosts other than localhost.

The Mongo connection is optional.  However, the `Listener` objects
mostly assume a database connection.  The `EABN` package, in
particular, has a mechanism for sending and recieving JSON messages
without using the database as a queue.

## R packages

The following R packages are used by the system:

* `jsonlite` -- Used for parsing JSON messages.
* `futile.logger` -- Used for logging system

The following R packages are suggested:

* `mongolite` -- Required if mongo database is used.
* `rmarkdown`, `knitr` -- Used to build vignettes.

# Configuring the Mongo Database


setupDatabases.js

## Proc 4 collections

## Authorized Apps

# Configuration files

## Configuration Directory 

/usr/local/share/Proc4

## Javascript configuration Files

Proc4.ini, Proc4.js

## R initialization files

EAini.R, EIini.R

# The PHP Dongle

## Game interface

PlayerStart.php
PlayerStop.php
PlayerStats.php
PlayerLevels.php


